The illusory promise of the Aligned Rank Transform

### 2025-12-12 text/10-intro.qmd
working on appendix I

This paper aims to clarify the severity of these issues and understand the potential risks of the method. We present the outcomes of a series of Monte Carlo experiments, following a distinctive simulation methodology grounded in latent variable modeling. This approach enables us to simulate effects consistently across a broad spectrum of distributions, both discrete and continuous, and allows us to address issues in interpreting main and interaction effects. To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).

Our findings corroborate Lüpsen's alarming conclusions. We provide overwhelming evidence that ART confounds effects and raises Type I error rates at very high levels across a diverse array of non-normal distributions, including skewed, binomial, and ordinal distributions, as well as distributions with unequal variances. These issues persist for both main and interaction effects. Our results further show that simpler rank transformation methods outperform ART, while parametric ANOVA generally poses fewer risks than ART when distributions deviate from normal. Given these new insights, we conclude that ART is not a viable analysis method and advocate for its abandonment. We provide recommendations for alternative analysis methods, while we also raise warnings about the interpretation of interaction effects.

>>>>

This paper aims to clarify the severity of these issues and understand the potential risks of the method. We present the results of a series of Monte Carlo experiments in which we simulate a broad spectrum of distributions, both discrete and continuous. Throughout our analysis, we pay special attention to definition-related issues in the null hypothesis, as these can affect the interpretation of the simulation results. To ensure the clarity of our findings and facilitate their reproducibility, we divide the experimental process into multiple smaller experiments, where each experiment focuses on a distinct variable (e.g., sample size, experimental design, variance ratio) or measure (Type I error rate, power, precision of effect size estimates).

Our findings corroborate Lüpsen's alarming conclusions. We provide overwhelming evidence that ART confounds effects and raises Type I error rates at very high levels across a diverse array of non-normal distributions, including skewed, binomial, and ordinal distributions, as well as distributions with unequal variances. These issues persist for both main and interaction effects. Our results further show that simpler rank transformation methods outperform ART, while parametric ANOVA generally poses fewer risks than ART when distributions deviate from normal. Given these new insights, we conclude that ART is not a viable analysis method and advocate for its abandonment. We provide recommendations for alternative analysis methods, while we also raise warnings about the analysis and interpretation of interaction effects.

### 2025-12-12 text/20-background.qmd
working on appendix I

Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology. In particular, we adopt a latent variable modeling approach that establishes a unified framework for data generation across all distributions. This approach allows us to identify interpretation issues regarding the definition of the null hypothesis, especially for interaction effects, and better explain when and why each method fails. Moreover, it facilitates the assessment of the methods through more suitable generation procedures for ordinal data, as suggested by @Liddell:2018.
Our simulation strategy is similar to that of @elkin:2021 but employs a simpler method for generating data from a latent parameter space. In addition, we evaluate critical scenarios that @elkin:2021 did not consider, particularly the influence of no-null effects and discrete data.

>>>>

Our goal is to verify Lüpsen's findings by employing an alternative set of experimental configurations and a distinctly different methodology based on generalized linear models. While our simulation strategy is similar to that of @elkin:2021, we investigate a larger range of experimental settings and critical scenarios the authors did not consider, particularly the influence of no-null effects and discrete data.

### 2025-12-12 text/40-interpretingEffects.qmd
working on appendix I

Statistical procedures based on GLMs define interaction in terms of the interaction coefficient $a_{12}$ in the linear component of the model. That is, the null hypothesis should be rejected when the coefficient is non-zero. However, there is no consensus among researchers on this approach. We clarify the debate and explain how we address it.

@@

**When the interpretation of effects becomes ambiguous.** Issues with interpreting interaction effects arise only when there are parallel main effects --- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe a change in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero. We discuss such scenarios in more depth in the [appendix](appendix.html#removable-main).

By contrast, interpretation is unambiguous in these situations:

@@

For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).

@@

### Our approach {#approach}
Responses in a generalized linear model can follow arbitrary distributions. A key challenge is how to simulate such distributions while maintaining control over the parameters of the linear submodel. We address this by adopting a latent variable modeling approach, in which a normally distributed latent variable is transformed into the observed response variable.

>>>>

Statistical procedures based on GLMs define effects in terms of the coefficients $a_{1}$, $a_{2}$, and $a_{12}$ in the linear component of the model. That is, the null hypothesis should be rejected when the corresponding coefficient (e.g., the coefficient $a_{12}$ for the interaction) is non-zero. However, there is no consensus among researchers on this approach. We clarify the debate and explain how we address it.

@@

Nevertheless, in most of our experiments we select scenarios that avoid these interpretational complications, ensuring that the Type I error rates of the methods we compare do not depend on how one defines interactions. We clarify this point below.

**Clarifying when the interpretation of effects becomes ambiguous.** Ambiguities in interpreting interaction effects arise only when all relevant main effects are present --- that is, when the coefficients of all interacting factors are non-zero. Interpretation issues may also extend to main effects when the interaction coefficient $a_{12}$ is non-zero. In such cases, we may observe changes in the marginal effect of a factor (e.g., $x_1$), even if its associated model coefficient (e.g., $a_1$) is zero. We discuss such scenarios in more detail in the [appendix](appendix.html#removable-main).

By contrast, interpretation is unambiguous in the following situations:

@@

In these cases, error rates (Type I or Type II) are comparable across all methods, regardless of whether the null hypothesis is defined in terms of the coefficients of the linear model or the observed effects on the response variable $Y$. For example, the divergence in ART's results in our illustrative example (see @fig-example) cannot be attributed to interpretation issues, since the sample was drawn from a population with no effect of *Technique* ($a_{1} = 0$) and no interaction ($a_{12}=0$).

@@

<!--
### Our approach {#approach}
Responses in a generalized linear model can follow arbitrary distributions. A key challenge is how to simulate such distributions while maintaining control over the parameters of the linear submodel. We address this by adopting a latent variable modeling approach, in which a normally distributed latent variable is transformed into the observed response variable.

### 2025-12-12 text/65-Results-EffectSize.qmd
working on appendix I

::: {#fig-eta-squared}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("data/effect-sizes-lognormal.csv", sep=",", header=TRUE, strip.white=TRUE)
data$effectX2 <- factor(data$effectX2)

@@

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

@@

::: {#fig-eta-squared-interaction}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("data/effect-sizes-normal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
data$effectX2 <- factor(data$effectX2)

@@

df <- data

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

@@

::: {#fig-eta-squared-ordinal}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("data/effect-sizes-ordinal-interactions.csv", sep=",", header=TRUE, strip.white=TRUE)
data$effectX2 <- factor(data$effectX2)

@@

df <- data
#df <- data[data$effectX1 == 0,]

fig1 <- plot_ly(df, x = ~effectX2, y = ~X1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

>>>>

::: {#fig-eta-squared}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("results/eta_squared.csv", sep=",", header=TRUE, strip.white=TRUE)
data <- data[data$family == "lnorm", ]
data$effectX2 <- factor(data$effectX2)

@@

fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

@@

::: {#fig-eta-squared-interaction}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
data <- read.csv("results/eta_squared.csv", sep=",", header=TRUE, strip.white=TRUE)
df <- data[data$family == "norm" & data$effectX1 == data$effectX2, ]
df$effectX2 <- factor(df$effectX2)

@@

fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))

@@

::: {#fig-eta-squared-ordinal}
```{r, echo=FALSE, warning=FALSE, fig.height=3.5, fig.width=8}
df <- data[data$family == "likert" & data$effectX1 == data$effectX2, ]
df$effectX2 <- factor(df$effectX2)

@@

fig1 <- plot_ly(df, x = ~effectX2, y = ~grtX1X2, type = 'box', boxpoints = 'all', name = 'ground truth', hovertemplate = " ", marker = list(color = cbPalette2[1]), fillcolor = cbPalette2[1], line = list(color = cbPalette[1]))  %>% layout(xaxis = list(title = NA))
fig.par <- plot_ly(df, x = ~effectX2, y = ~parX1X2, type = 'box', boxpoints = 'all', name = 'PAR', hovertemplate = " ", marker = list(color = cbPalette2[2]), fillcolor = cbPalette2[2], line = list(color = cbPalette[2])) %>% layout(xaxis = list(title = NA))
fig.rnk <- plot_ly(df, x = ~effectX2, y = ~rnkX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'RNK', marker = list(color = cbPalette2[3]), fillcolor = cbPalette2[3], line = list(color = cbPalette[3])) %>% layout(xaxis = list(title = NA))
fig.int <- plot_ly(df, x = ~effectX2, y = ~intX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'INT', marker = list(color = cbPalette2[4]), fillcolor = cbPalette2[4], line = list(color = cbPalette[4])) %>% layout(xaxis = list(title = NA))
fig.art <- plot_ly(df, x = ~effectX2, y = ~artX1X2, type = 'box', boxpoints = 'all', hovertemplate = " ", name = 'ART', marker = list(color = cbPalette2[5]), fillcolor = cbPalette2[5], line = list(color = cbPalette[5])) %>% layout(xaxis = list(title = NA))
